Basic Model formulation

### 2015-02-27 documentation/model_formulation.Rmd
update model formulation, rely on simfunctions.r; add mean field model

parms <- list(
    m = 0.05,
    r = 1,
    b = 0.9, 
    K = 0.9, 
    a = 0.6,
    h = 100,
    L = 5, 
    alpha = 0,
    q = 0,
    c = 0,
    f = 0,
    v = 0,
    p = 0
  )

par(mar = c(4,4,2,4))
plot(rho,  with(parms,  r * b * rho * ( 1 - (rho / K)) / (1 - rho) ) , ylab = "individual growth on empty cells", type = "l", ylim = c(0,2.5), las = 1, yaxs = "i")

@@

parms <- list(
    m = 0.05,
    r = 1,
    b = 0.9, 
    K = 0.9, 
    a = 0.6,
    h = 100,
    L = 16, 
    alpha = 0.1,
    q = 1,
    c = 0,
    f = 0,
    v = 0,
    p = 0

  )

@@

G <- function(V, q_01 = "auto", r_0 = 1, b = 0.1, f = 0, alpha = 0,  K = 0.9, c = 0) {
  if(q_01 == "auto") q_01 <- 1
  r_0*(b + (1-b)*f*q_01)*V^(1+alpha)*(1-V/(K * (1-c * q_01) ) )

}

defplot(rho, G(rho, 1, f = 0.8), ylab = "growth", col = "green3")
lines(rho,G(rho, 0.75, f = 0.8), col = "green3")
lines(rho,G(rho, 0.5, f = 0.8), col = "green3")
lines(rho,G(rho, 0.25, f = 0.8), col = "green3")
lines(rho,G(rho, 0, f = 0.8), col = "green3")

lines(rho, C(rho, L = 10))

@@

defplot(rho, G(rho, 1, c = 0.3, b = 0.9), ylab = "growth", col = "green3")
lines(rho,G(rho, 0.75, c = 0.3, b = 0.9), col = "green3")
lines(rho,G(rho, 0.5, c = 0.3, b = 0.9), col = "green3")
lines(rho,G(rho, 0.25, c = 0.3, b = 0.9), col = "green3")
lines(rho,G(rho, 0, c = 0.3, b = 0.9), col = "green3")

lines(rho, C(rho, L = 10))

@@

C <- function(V, q_01 = 1, m_0 = 0.05, L = 5, a = 0.1, h = 100, q = 0, p = 0, v = 0) (m_0 * V) +( (a+v*q_01  ) *V^(1+q)*L*(1-p*q_01))/(1+(a+v*q_01)*h*(V^(1+q)) )

defplot(rho, C(rho, 1, v = 0.5), ylab = "mortality")
lines(rho,C(rho, 0.75, v = 0.5))
lines(rho,C(rho, 0.5, v = 0.5))
lines(rho,C(rho, 0.25, v = 0.5))
lines(rho,C(rho, 0, v = 0.5))

lines(rho, G(rho, b = 0.6), col = "green3")

@@

defplot(rho, C(rho, 1, p = 0.8, L = 10), ylab = "mortality")
lines(rho,C(rho, 0.75, p = 0.8, L = 10))
lines(rho,C(rho, 0.5, p = 0.8, L = 10))
lines(rho,C(rho, 0.25, p = 0.8, L = 10))
lines(rho,C(rho, 0, p = 0.8, L = 10))

lines(rho, G(rho, b = 0.6), col = "green3")

@@

defplot(rho, C(rho, 1, p = 0.8, v = 0.6, L = 10, q = .5), ylab = "mortality")
lines(rho,C(rho, 0.75, p = 0.8, v = 0.6, L = 10, q = .5))
lines(rho,C(rho, 0.5, p = 0.8,v = 0.6,  L = 10, q = .5))
lines(rho,C(rho, 0.25, p = 0.8,v = 0.6,  L = 10, q = .5))
lines(rho,C(rho, 0, p = 0.8,v = 0.6,  L = 10, q = .5))

lines(rho, G(rho, 1, c = 0.3, f = 0.9, b = 0.3, alpha = 0.5), col = "green3")
lines(rho,G(rho, 0.75, c = 0.3,f = 0.9,  b = 0.3, alpha = 0.5), col = "green3")
lines(rho,G(rho, 0.5, c = 0.3,f = 0.9,  b = 0.3, alpha = 0.5), col = "green3")
lines(rho,G(rho, 0.25, c = 0.3, f = 0.9, b = 0.3, alpha = 0.5), col = "green3")
lines(rho,G(rho, 0, c = 0.3,f = 0.9,  b = 0.3, alpha = 0.5), col = "green3")

@@

@@

ini_rho <- function(rho_1, rho_11 = NULL) {
    if(is.null(rho_11)) out <- {
      c(
    rho_1 = rho_1,
    rho_11 = rho_1*rho_1,
    rho_10 = rho_1*(1-rho_1)*2,
    rho_00 = (1-rho_1)*(1-rho_1),
    rho_0 = 1-rho_1
    ) 
    } else {
      out <- c(
    rho_1 = rho_1,
    rho_11 = rho_11,
    rho_10 = rho_1-rho_11,
    rho_00 = 1-2*rho_1+rho_11,
    rho_0 = 1-rho_1
    )
    if(any(out < 0)) out <- c(
    rho_1 = NA,
    rho_11 = NA,
    rho_10 = NA,
    rho_00 = NA,
    rho_0 = NA
    )
    }

  return(out)
}

G <- function(rho_1, q_01 = "auto", prms = parms) {
  if(q_01 == "auto") q_01 <- 1
  with(prms, r*(b + (1-b)*f*q_01)*rho_1^(1+alpha)*(1-rho_1/(K * (1-c * q_01) ) ))

}

C <- function(rho_1, q_11 = 1, prms = parms) with(prms, (m * rho_1) +( (a+v*q_11  ) *rho_1^(1+q)*L*(1-p*q_11))/(1+(a+v*q_11)*h*(rho_1^(1+q)) ) )

attractor <- function(parms, rho_1_ini = seq(0,1, length = 41), rho_11_ini = seq(0,1, length = 11)) {

ini <- list(
  rho_1 = rho_1_ini,
  rho_11 = rho_11_ini,
  rho_10 = NA,
  rho_00 = NA,
  rho_0 = NA
  )

ini <- expand.grid(ini)

for(x in 1:nrow(ini)) {
      temp <- ini_rho(ini[x,]$rho_1, ini[x,]$rho_11)
      ini[x,]$rho_1 <- temp[1]
      ini[x,]$rho_11 <- temp[2]
      ini[x,]$rho_10 <- temp[3]
      ini[x,]$rho_00 <- temp[4]
      ini[x,]$rho_0 <- temp[5]
      }

ini$m_ini <- C(ini$rho_1, ini$rho_11/ini$rho_1)
ini$g_ini <- G(ini$rho_1, (ini$rho_1-ini$rho_11)/(1-ini$rho_1))

ini <- subset(ini, !is.na(m_ini))
ini <- cbind(ID = 1:nrow(ini),ini)

foreach(iteration = ini$ID, .combine = rbind, .packages = c("deSolve")) %dopar% { 
  source("C:/Users/SCHNEIDER/Documents/projects/CAS02_livestock/code/simfunctions.r")

   rho_starting <- ini[iteration, 2:6]

  # running the ode-solver
  runmodel <- ode(y = as.numeric(rho_starting), func = odesys_spex, times = seq(1,2), parms = parms)

  out <-as.data.frame(runmodel)[dim(runmodel)[1],]

  # transfer into ouput and calculate missing rho values
  return(out)
  } -> output

names(output) <- c("time", "rho_1", "rho_11", "rho_10", "rho_00", "rho_0")

defplot(rho, C(rho, 1), ylab = "mortality", col= "#00000020")
lines(rho,C(rho, 0.75), col= "#00000040")
lines(rho,C(rho, 0.5), col= "#00000080")
lines(rho,C(rho, 0.25), col= "#00000040")
lines(rho,C(rho, 0), col= "#00000020")

lines(rho, G(rho, 1), col = "#00993320")
lines(rho,G(rho, 0.75), col = "#00993340")
lines(rho,G(rho, 0.5), col = "#00993380")
lines(rho,G(rho, 0.25), col = "#00993340")
lines(rho,G(rho, 0), col = "#00993320")

arrows(ini$rho_1, C(ini$rho_1, ini$rho_11/ini$rho_1),output$rho_1, C(output$rho_1, output$rho_11/output$rho_1), length = 0.02)

arrows(ini$rho_1, G(ini$rho_1, (ini$rho_1-ini$rho_11)/(1-ini$rho_1)),output$rho_1, G(output$rho_1, (output$rho_1-output$rho_11)/(1-output$rho_1) ), length = 0.02, col = "#009933" )

high_equ <- as.data.frame(ode(y = ini_rho(0.8), func = odesys_spex, times = c(1,1000), parms = parms) )[2,]

points(high_equ$rho_1, C(high_equ$rho_1, high_equ$rho_11/high_equ$rho_1), pch = 20)

low_equ <- as.data.frame(ode(y = ini_rho(0.0001), func = odesys_spex, times = c(1,1000), parms = parms) )[2,]

points(low_equ$rho_1, C(low_equ$rho_1,low_equ$rho_11/low_equ$rho_1), pch = 20)

}

>>>>

defparms <- list(
    m = 0.05,
    r = 1,
    b = 0.9, 
    K = 0.9, 
    a = 0.6,
    h = 100,
    L = 5, 
    alpha = 0,
    q = 0,
    c = 0,
    f = 0,
    v = 0,
    p = 0
  )
parms <- defparms
par(mar = c(4,4,2,4))
plot(rho,  with(parms,  r * b * rho * ( 1 - (rho / K)) / (1 - rho) ) , ylab = "individual growth on empty cells", type = "l", ylim = c(0,2.5), las = 1, yaxs = "i")

@@

defparms <- list(
    m = 0.05,
    r = 1,
    b = 0.9, 
    K = 0.9, 
    a = 0.6,
    h = 100,
    L = 16, 
    alpha = 0.1,
    q = 1,
    c = 0,
    f = 0,
    v = 0,
    p = 0

  )
parms <- defparms

@@

defplot(rho, G(rho, 1, set = list( f = 0.8)), ylab = "growth", col = "green3")
lines(rho,G(rho, 0.75, set = list( f = 0.8)), col = "green3")
lines(rho,G(rho, 0.5, set = list( f = 0.8)), col = "green3")
lines(rho,G(rho, 0.25, set = list( f = 0.8)), col = "green3")
lines(rho,G(rho, 0, set = list( f = 0.8)), col = "green3")

lines(rho, C(rho, set = list( L = 10)))

@@

defplot(rho, G(rho, 1, set= list(c = 0.3, b = 0.9)), ylab = "growth", col = "green3")
lines(rho,G(rho, 0.75, set= list( c = 0.3, b = 0.9)), col = "green3")
lines(rho,G(rho, 0.5, set= list( c = 0.3, b = 0.9)), col = "green3")
lines(rho,G(rho, 0.25, set= list(c = 0.3, b = 0.9)), col = "green3")
lines(rho,G(rho, 0, set= list( c = 0.3, b = 0.9)), col = "green3")

lines(rho, C(rho, set= list( L = 10)))

@@

defplot(rho, C(rho, 1, set = list( v = 0.5)), ylab = "mortality")
lines(rho,C(rho, 0.75, set = list( v = 0.5)))
lines(rho,C(rho, 0.5, set = list( v = 0.5)))
lines(rho,C(rho, 0.25, set = list( v = 0.5)))
lines(rho,C(rho, 0, set = list( v = 0.5)))

lines(rho, G(rho, set = list( b = 0.6)), col = "green3")

@@

defplot(rho, C(rho, 1, set= list(p = 0.8, L = 10)), ylab = "mortality")
lines(rho,C(rho, 0.75, set= list(p = 0.8, L = 10)))
lines(rho,C(rho, 0.5, set= list(p = 0.8, L = 10)))
lines(rho,C(rho, 0.25, set= list(p = 0.8, L = 10)))
lines(rho,C(rho, 0, set= list(p = 0.8, L = 10)))

lines(rho, G(rho, set= list(b = 0.6)), col = "green3")

@@

defplot(rho, C(rho, 1, set = list( p = 0.8, v = 0.6, L = 10, q = .5)), ylab = "mortality")
lines(rho,C(rho, 0.75, set = list( p = 0.8, v = 0.6, L = 10, q = .5)))
lines(rho,C(rho, 0.5, set = list( p = 0.8,v = 0.6,  L = 10, q = .5)))
lines(rho,C(rho, 0.25, set = list( p = 0.8,v = 0.6,  L = 10, q = .5)))
lines(rho,C(rho, 0, set = list( p = 0.8,v = 0.6,  L = 10, q = .5)))

lines(rho, G(rho, 1, set = list(c = 0.3, f = 0.9, b = 0.3, alpha = 0.5)), col = "green3")
lines(rho,G(rho, 0.75, set = list( c = 0.3,f = 0.9,  b = 0.3, alpha = 0.5)), col = "green3")
lines(rho,G(rho, 0.5, set = list(c = 0.3,f = 0.9,  b = 0.3, alpha = 0.5)), col = "green3")
lines(rho,G(rho, 0.25, set = list( c = 0.3, f = 0.9, b = 0.3, alpha = 0.5)), col = "green3")
lines(rho,G(rho, 0, set = list(c = 0.3,f = 0.9,  b = 0.3, alpha = 0.5)), col = "green3")

@@

### mean field model

As a simple approximation of the model serves a mean field model, assuming that the conditional local vegetation cover of empty cells $q_{0|1}$ and of vegetated cells $q_{1|1}$ both are well represented by the global vegetation cover $\rho_q$. Of course $\rho_0 = 1-\rho_1$.
The change in cover derives from the growth and mortality terms

Thus, we can substitute $w_{1,0}$ and $w_{0,1}$ and obtain

$$  \frac{\mathrm{d}\rho_{1}}{\mathrm{d}t} = r \rho_1^{1+\alpha} (b + (1 - b)  f \rho_1)) (1-\frac{\rho_1}{K(1 - c \rho_1}) - \rho_{1}m - \frac{(a  + v \rho_1)L (1 - p \rho_1)\rho_1^{1+q}}{1+(a  + v \rho_1)h\rho_1^{1+q}}  $$

Plotting the meanfield growth and mortality over vegetation cover is straigthforward now.

$$ G = r \rho_1^{1+\alpha} (b + (1 - b)  f \rho_1)) (1-\frac{\rho_1}{K(1 - c \rho_1}) $$

$$ C = \rho_{1}m + \frac{(a  + v \rho_1)L (1 - p \rho_1)\rho_1^{1+q}}{1+(a  + v \rho_1)h\rho_1^{1+q}}$$

The steady states can be simulated by solving the ODE.

parms <- defparms
parms$f = 0.9
parms$b = 0.4
parms$L = 5

attractor(parms, pairapprox = FALSE, localvals = TRUE)

But the steady state also can be solved analytically.

Therefore, we solve for the equilibrium case

$$  \frac{\mathrm{d}\rho_{1}}{\mathrm{d}t} = 0 $$

$$  0 = r \rho_1^{1+\alpha} (b + (1 - b)  f \rho_1)) (1-\frac{\rho_1}{K(1 - c \rho_1}) - \rho_{1}m - \frac{(a  + v \rho_1)L (1 - p \rho_1)\rho_1^{1+q}}{1+(a  + v \rho_1)h\rho_1^{1+q}}  $$

@@
